\documentclass[a4paper]{article}

\def\npart{II}
\def\ntitle{Integrable Systems}
\def\nlecturer{C.\ Warnick}

\def\nterm{Michaelmas}
\def\nyear{2017}

\input{header}

\begin{document}

\input{titlepage}

Consider
\[
  u_t + u_{xxx} =0
\]
Seek a solutioin of the form
\[
  u(x,t) = \exp(-i\omega t+ ikx)
\]
This solve the equation if
\[
  \omega = -k^3
\]
dispersion relation
so
\[
  u_k(x,t) = \exp(ik(k^t+x)
\]

Each \(u_k\) is a wave translating at speed \(k^2\) in negative \(x\)-direction. Since speed depends on frequency, we say the equation is \emph{dispersive}. We can consider a superposition of solutions
\[
  u(x,t) = \int_{\R}^{ } a(k)\exp(ik(k^2t+x) dk
\]
where \(a(k)\) is arbitrary and is related to \(\hat{u(\cdot,0)}\). Since different frequencies move at different speeds, solutions should spread out over time.

\subsection{Nonlinear Growth}

Consider
\[
  u_t - 6 uu_x =0,
\]
the \emph{Burgers equation}. It can be solved by the method of characteristics to give
\[
  u(x,t) = f(\xi), \zeta = x+6tf(\xi).
\]
and
\[
  f(\xi) = u(\xi,0).
\]

It can be shown that this solution becomes singular, in the sense that \(\sup_{x\in\R}|U_x(x,t)| \to \infty\) as \(t\to \infty\) where
\[
  T = \inf_{x\in\R} \frac{1}{6f'(x)}.
\]
Surprisingly, the dispersive and nonlinear growth effects balance for KdV. It was known for some time that KdV admits a soliton solution
\[
  u(x,t) = -2x^2 \sech^2 \chi (x-2x^4t-x_0)
\]
solves KdV for \(x,x_0\in \R\). It represents a wave moving to the right with speed \(4x^2\).

In the mid 60s, Kruskal and Zabosky simulated a solution that starts as two solitons, set up so that one overtakes the other. Curiously, these solitons passed through one another without interacting. If KdV were linear, this would be a surprise, but it is not.

\subsection{Sine-Gordon Equation}

The Sine-Gordon equation is the PDE
\[
  u_{tt} - u_{xx} + \delta uu = 0.
\]
This is the equation modelling an infinite series of pendula connected by springs.

Sine-Gordon also admits soliton solutions. The \(1\)-soliton solution is given by
\[
  u(x,t) = 4\tan^{-1} \exp \Big( \frac{x-vt}{\sqrt{1-v^2}} \Big)
\]
This is a ``kink'' solution. The wave translates uniformly with speed \(v\). As for KdV, soliton solutions pass through one another without interaction.

\subsection{Backlund Transformations}

Recall the Cauchy-Riemann equations
\begin{align*}
  u_x &= \sqrt y \\
  -u_y &= \sqrt x
\end{align*}
so that
\[
  u_{xx} + u_{yy} = 0.
\]

If we are given \(v\) a solution of Laplace, we can find \(u\) such that
\begin{align*}
  u_x &= \sqrt y \\
  u_y &= -\sqrt x
\end{align*}
Then \(u\) will automatically solve Laplace.

\begin{eg}
  Let \(v=xy\). Solve
  \begin{align*}
    u_x &= x \\
    u_y &= -y 
  \end{align*}
This is an exmaple of a (auto) Backlund transformation, relating one solution to another by first order system.
\end{eg}

For a more interesting example, take Sine-Gordon in light cone coordinates
\[
  \frac{\p^2 u}{\p\xi \p\tau} = \sin u.
\]
Consider the equations
\begin{align*}
  \p_\xi(\phi_1-\phi_0) &= 2\varepsilon \sin \Big( \frac{\phi_1+\phi_0}{2} \Big) \\
  \p_\tau(\phi_1-\phi_0) &= \frac{2}{\varepsilon} \sin \Big( \frac{\phi_1-\phi_0}{2} \Big)
\end{align*}
Differentiating the first equation with repect to \(\tau\):
\begin{align*}
  \p_\psi\p_\xi(\phi_1-\phi_0) &= 2\varepsilon \p_\tau \frac{\phi_1+\phi_0}{2}\cos \Big( \frac{\phi_1+\phi_0}{2} \Big) \\
                               &= 2\sin\frac{\phi_1-\phi_0}{2} \cos\frac{\phi_1+\phi_0}{2} \\
                               &= \sin\phi_1 - \sin\phi_0
\end{align*}
so if \(\phi_0\) sovles Sine-Gordon then \(\phi_1\) must as well

\begin{ex}
  If \(\psi_0 = 0\), we recover the \(1\)-soliton solution.
\end{ex}
\end{document}